# Neighborhood Sample

## Data

Data come from two sources. The first are variables downloaded from various city sources for the number of vacant buildings, the number of 311 requests for street repairs, and the median (?) date of housing construction. These variables are saved in the CSV file `data/potholes-vacants-yearbuilt.csv`. 

The last variable is the number of housing units. This variable comes from table `B25001` in the American Community Survey. That data is obtained using the `get_acs()` function of the `tidycensus` library, which also downloads geography to be used for mapping values. 

I merge the two tables and then create a variable that calculates the ratio of vacant buildings to housing units in the tract (this is a rough adjustment for number of buildings). 

```{r data, message=FALSE, results='hide'}
## Data from City Sources
dta <- read_csv("data/potholes-vacants-yearbuilt.csv", col_types = "cciii") %>% 
    mutate(
        
    ) %>% 
    rename(GEOID = tract) 

## ACS Data (including geography)
hu <- get_acs(
    geography = "tract",
    variables = "B25001_001",
    state = "MD",
    county = "510",
    geometry = TRUE
) %>% 
    sf::st_transform(crs = sf::st_crs("EPSG:6487")) %>% 
    rename(hu = estimate) %>% 
    select(GEOID, hu) %>% 
    left_join(dta, by = "GEOID") %>% 
    mutate(
        lnvacant = log(vacant + 1),
        vacrat = vacant / hu,
        lnvacrat = if_else(hu > 0, log((vacant + 1) / hu), NA_real_)
    )
```

## Univariate Distributions

Density the univariate distributions of the three variables we are examining for this analysis.

```{r univariate, fig.asp=1/3}
## Specify variables and labels
varnms <- c("vacrat", "streetrepairs", "yearblt")
varlab <- c("Vacant ratio", "Street repair requests", "Year built")
names(varlab) <- varnms

## Set unique limits and breaks for plots (also used in maps below)
varlims <- list(
    vacrat = c(0, 0.55),
    streetrepairs = c(0, 3100),
    yearblt = c(1890, 2010)
)

varbrks <- list(
    vacrat = seq(0, 0.55, 0.1),
    streetrepairs = seq(0, 3000, 1000),
    yearblt = seq(1890, 2010, 30)
)

varlabs <- list(
    vacrat = sprintf("%.1f", varbrks[[1]]), 
    streetrepairs = as.character(varbrks[[2]]),
    yearblt = as.character(varbrks[[3]])
)

## Function to create univariate density plot
densplot <- function(v) {
    ggplot(hu, aes(x = .data[[v]])) +
        geom_density() +
        scale_x_continuous(
            limits = varlims[[v]], breaks = varbrks[[v]]
        ) +
        labs(x = NULL, y = NULL)
}

## Plot density plots into row
d_plts <- map(varnms, densplot)
d_pltrow <- ggarrange(d_plts[[1]], d_plts[[2]], d_plts[[3]], nrow = 1)
d_pltrow
```

## Bivariate/Trivariate Relationships

Calculate the correlations between the variables. 

```{r correlation-matrix}
hu %>% 
    as_tibble() %>% 
    select(lnvacrat, streetrepairs, yearblt) %>% 
    as.matrix() %>% 
    cor(use = "pairwise") %>% 
    round(2) 
```

Plot bivariate scatterplots of the three variables and a correlation matrix. 

```{r bivariate, warning=FALSE, message=FALSE}
sct1 <- ggplot(hu, aes(x = streetrepairs, y = yearblt)) + 
    geom_point(size = 0.75) + geom_smooth(method = "lm", se = FALSE)
sct2 <- ggplot(hu, aes(x = streetrepairs, y = lnvacrat)) +
    geom_point(size = 0.75) + geom_smooth(method = "lm", se = FALSE)
sct3 <- ggplot(hu, aes(x = yearblt, y = lnvacrat)) +
    geom_point(size = 0.75) + geom_smooth(method = "lm", se = FALSE)
ttl <- ggplot() + labs(title = "Bivariate relationships among variables") +
    theme(title = element_text(face = "bold"))

cor_plt <- ggarrange(
    ggarrange(ttl),
    ggarrange(sct1, NULL, sct2, sct3),
    nrow = 2, heights = c(.1, .9)
)
cor_plt

ggsave('out/img/scatterplots.pdf', width = 7.25, height = 5.5, units = "in")
```

## Map Distributions

Below, we plot values of the three variables onto maps of Baltimore. The white boundaries represent the Census public-use microdata areas (PUMAs). The figure below is saved to `out/img/variable-maps-densities.pdf`.

```{r maps, results='hide'}
## Download PUMA geography
pumas <- pumas(state = "24", year = 2022) %>% 
    filter(PUMACE20 %in% paste0("0080", 1:5))

## Download and crop Chesapeake Bay geography
f_chpk <- "data/Chesapeake_Bay_92_Segments/Chesapeake_Bay_92_Segments.shp"
dta_chpk <- st_read(f_chpk) %>% 
    sf::st_transform(crs = sf::st_crs("EPSG:6487")) %>% 
    st_crop(st_bbox(hu))


bmplot <- function(v) {
    ggplot() +
        geom_sf(data = hu, aes(fill = .data[[v]])) +
        geom_sf(data = dta_chpk, fill = "white", color = "#dddddd", linewidth = .35) +
        geom_sf(data = pumas, alpha = 0, color = "white", linewidth = .5) +
        scale_fill_viridis_c(
            limits = varlims[[v]], breaks = varbrks[[v]], 
            labels = varlabs[[v]]
        )+
        labs(
            title = varlab[v],
            fill = NULL
        ) +
        theme_map
}

bm_plts <- map(varnms, bmplot)
ttl <- ggplot() + labs(title = "Distributions of built environment variables") +
    theme(title = element_text(face = "bold"))

d_plts_c <- map(d_plts, function(p) {
    p + theme(
        axis.text.y = element_blank()
    )
})

ggarrange(
    ttl,
    ggarrange(bm_plts[[1]], bm_plts[[2]], bm_plts[[3]], nrow = 1), 
    ggarrange(d_plts_c[[1]], d_plts_c[[2]], d_plts_c[[3]], nrow = 1), 
    nrow = 3, heights = c(.075, .45, .15)
)

ggsave("out/img/variable-maps-densities.pdf", width = 7.25, height = 5.5, units = "in")

```

## Proposed Sample Neighborhoods

Based on variation across the three different variables above (and our knowledge of the racial/ethnic and socioeconomic distributions), we may consider sampling the following neighborhoods. The neighborhood names come from [spatial data][nhd-data] maintained by Baltimore City official list.

[nhd-data]: https://data.baltimorecity.gov/datasets/47f7b87abafa487595e1f837135b0ed1/explore

```{r proposed-nhds, echo=TRUE}
proposed_nhds <- list(
    # Central Baltimore
    central = c(
    "Old Goucher", "Barclay", "East Baltimore Midway", "Broadway East",
    "Berea"
    ), 
    # Northwest
    northwest = c(
    "Fallstaff", "Cross Country", "Glen", "Cheswolde"
    ),
    # North Central
    north_central = c(
    "North Roland Park/Poplar Hill", "The Orchards", "Bellona-Gittings", 
    "Cedarcroft", "Rosebank", "Lake Evesham", "Homeland"
    ),
    # West
    west = c(
    "Carrollton Ridge", "Boyd-Booth", "Penrose/Fayette Street Outreach",
    "Evergreen Lawn", "Bridgeview/Greenlawn", "Rosemont Homeowners/Tenants",
    "Midtown-Edmondson"
    ),
    # South
    south = c(
    "Cherry Hill", "Brooklyn"
    ),
    # East
    east = c(
    "Highlandtown", "Greektown", "Canton", "Brewers Hill"
    )
)
```

The proposed neighborhoods are plotted in orange below. This covers variation across our three variables and some racial and socioeconomic variation. It does, however, miss much of East Baltimore (especially the Northeast). Mac and I ran an experiment a while back and covered the streets along the route [at this link][strava] in one hour and fifteen minutes. Based on that, it's possible that we can do the cluster of neighborhoods in the Northwest and in the Roland Park/Homeland area in one day, the Midtown and Old Goucher/Broadway East stretch another day, South Baltimore a third day, and have another day to do more on the East Side. 

[strava]: https://www.strava.com/activities/9303580510

```{r map-proposed, results='hide'}
f_nhd <- "data/Neighborhood/Neighborhood.shp"
dta_nhd <- st_read(f_nhd, quiet = TRUE)
dta_bal <- dta_nhd %>% 
  st_buffer(5) %>% 
  st_union() %>% 
  st_buffer(-5)
dta_pnhd <- dta_nhd %>% 
    filter(Name %in% unlist(proposed_nhds))
nrow(dta_pnhd) == length(proposed_nhds) # Check that all proposed neighborhoods
                                          # are in data

# Add proposed neighborhood to map plots
p_pnhd <- geom_sf(data = dta_pnhd, alpha = 0, color = "orange", linewidth = 0.65) 
p_build <- geom_sf(data = filter(dta_nhd, Name %in% c("Greenmount West", "Oliver")), fill = "red")
p_parity <- geom_sf(data = filter(dta_nhd, Name %in% c("Harlem Park")), fill = "lightblue")
p_coppin <- geom_sf(data = filter(dta_nhd, str_detect(Name, "Coppin")), fill = "pink")
bm_pnhds <- map(bm_plts, function(p) {return(p + p_pnhd + p_build + p_parity + p_coppin)})

ttl <- ggplot() + labs(title = "Proposed neighborhoods for Trimble sample") +
    theme(title = element_text(face = "bold"))

ggarrange(
    ttl,
    ggarrange(bm_pnhds[[1]], bm_pnhds[[2]], bm_pnhds[[3]], nrow = 1), 
    nrow = 2, heights = c(.039, .236)
)

ggsave(
    "out/img/proposed-sample-locations.pdf",
    width = 7.25, height = 5.5, units = "in"
)

```
Create a dataframe containing sampled areas that dissolves neighborhood boundaries. 

```{r smp-dissolve-geographies}
pnhdlabs <- unlist(proposed_nhds)
names(pnhdlabs) <- str_replace(names(pnhdlabs), "\\d+$", "")
dta_smpreg <- dta_pnhd %>% 
    mutate(
        region = fct_recode(Name, !!!pnhdlabs)
    ) %>% 
    st_buffer(100) %>% ## Avoid slivers by buffering by 500ft. 
    group_by(region) %>% 
    summarize(
    ) %>% 
    st_zm() 
```

### Reference Maps

The following maps provide a reference for the race/ethnicity and poverty composition of city neighborhoods. 

### Race/Ethnicity

```{r, race-ethnicity, message=FALSE}
dta_re <- get_acs(
    geography = "tract",
    table = "B03002",
    output = "wide",
    state = "MD",
    county = "510",
    geometry = TRUE,
    cache_table = TRUE
) %>% 
    rename_with(~str_replace(., "E$", ""))%>% 
    mutate(
        totpop = B03002_001,
        pnhw = B03002_003 / totpop,
        pnhb = B03002_004 / totpop,
        poth = (B03002_002 - (B03002_003 + B03002_004)) / totpop,
        phsp = B03002_012 / totpop
    ) %>% 
    sf::st_transform(crs = sf::st_crs("EPSG:6487"))


plot_re <- function(v, l) {
    ggplot() +
        geom_sf(data = dta_re, fill = "#bbbbbb") +
        geom_sf(data = filter(dta_re, totpop > 200), aes(fill = .data[[v]])) +
        geom_sf(data = dta_chpk, fill = "white", color = "#dddddd", linewidth = .35) +
        geom_sf(data = pumas, alpha = 0, color = "white", linewidth = .5) +
        scale_fill_distiller(
            palette = "Blues", direction = 1, limits = c(0, 1)
        ) +
        labs(
            title = paste("Percent", l)
        ) +
        theme_map
}

vars_re <- c(Black = "pnhb", White = "pnhw", Latinx = "phsp")
plts_re <- map2(vars_re, names(vars_re), plot_re)

ggarrange(
    plts_re[[1]], plts_re[[2]], plts_re[[3]],
    nrow = 1
)
```

### Poverty

```{r, poverty, message=FALSE}
dta_pov <- get_acs(
    geography = "tract",
    variables = c("B17001_001", "B17001_002"),
    output = "wide",
    state = "MD",
    county = "510",
    geometry = TRUE,
    cache_table = TRUE
) %>% 
    mutate(
        totpop = B17001_001E,
        ppov = B17001_002E / totpop,
        ppovc = cut(ppov, seq(0, .4, .05))
    )

ggplot() +
    geom_sf(data = dta_pov, fill = "#bbbbbb") +
    geom_sf(data = filter(dta_pov, totpop > 200), aes(fill = ppov)) +
    geom_sf(data = dta_chpk, fill = "white", color = "#dddddd", linewidth = .35) +
    geom_sf(data = pumas, alpha = 0, color = "white", linewidth = .5) +
    scale_fill_distiller(
        palette = "Blues", direction = 1, limits = c(0, .6),
        breaks = seq(0, 0.6, .1)
    ) +
    labs(
        title = "Percent poor",
        fill = NULL
    ) +
    theme_minimal() +
    theme(
        title = element_text(face = "bold"),
        legend.position = "top", 
        panel.grid = element_blank(),
        axis.text = element_blank(),
        legend.key.width = unit(2, "lines"),
        legend.margin = margin(b = -3, unit = "pt")
    )
# +
#     theme(
#         title = element_text(margin = margin(0, 0, 0, 0)),
#         legend.position = "top",
#         legend.margin = margin(0, 0, 0, 0)
#     )

```

```{r mapview}
## Use mapview to look have tooltip reveal census tract number
# library(mapview)
# mapview(hu, zcol = "vacrat", label = "GEOID")
# mapview(pumas, linewidth = 20, color = "white") + mapview(dta_nhd, label = "Name", alpha = .5) 
# mapview(filter(dta_nhd, Name %in% c("Oliver", "Greenmount West"))) 

```
